---
title: "Draft"
author: "Samaher"
date: "2024-02-18"
output: html_document
---

## 1. Introduction 

## 2. Data Used 

`tourney_id`: a unique identifier for each tournament.

`tourney_name`: the name of the tournament

`surface`: Hard, Clay, or Grass

`draw_size`: number of players in the draw, often rounded up to the nearest power of 2. (For instance, a tournament with 28 players may be shown as 32.)

`tourney_level`:
'G' = Grand Slams
'F' = WTA finals 
'PM' = Premier Mandatory
'P' = Premier
'I' = International
'D' is used for Federation/Fed/Billie Jean King Cup, and also for Wightman Cup and Bonne Bell Cup.

`tourney_date`: eight digits, YYYYMMDD, usually the Monday of the tournament week.

`match_num`: a match-specific identifier. Often starting from 1, sometimes counting down from 300, and sometimes arbitrary. 

`winner_id`: the player_id used in this repo for the winner of the match

`winner_seed`

`winner_entry`:
'WC' = wild card
'Q' = qualifier
'LL' = lucky loser
'ALT' = Alternate 
'SE' = Special Exemption
'SR' = Special Exempt

`winner_name`

`winner_hand`: R = right, L = left, U = unknown. For ambidextrous players, this is their serving hand.

`winner_ht`: height in centimeters, where available

`winner_ioc`: three-character country code

`winner_age`: age, in years, as of the tourney_date

`loser_id`

`loser_seed`

`loser_entry`

`loser_name`

`loser_hand`

`loser_ht`

`loser_ioc`

`loser_age`

`score`

`best_of`:'3', indicating the the number of sets for this match

`round`

`minutes`: match length, where available

`w_ace`: winner's number of aces

`w_df`: winner's number of doubles faults

`w_svpt`: winner's number of serve points

`w_1stIn`: winner's number of first serves made

`w_1stWon`: winner's number of first-serve points won

`w_2ndWon`: winner's number of second-serve points won

`w_SvGms`: winner's number of serve games

`w_bpSaved`: winner's number of break points saved

`w_bpFaced`: winner's number of break points faced

`l_ace`

`l_df`

`l_svpt`

`l_1stIn`

`l_1stWon`

`l_2ndWon`

`l_SvGms`

`l_bpSaved`

`l_bpFaced`

`winner_rank`: winner's WTA rank, as of the tourney_date, or the most recent ranking date before the tourney_date

`winner_rank_points`: number of ranking points, where available

`loser_rank`

`loser_rank_points`


```{r}
data <- read.csv("/Users/samaherbrahem/Documents/MSc DSE/Trimester 2/Statistical Learning/Individual Project/WTA/wta_matches_2023.csv")
```

## 3. Data Processing

```{r}
clean_data <- data[, !(names(data) %in% c("tourney_id", "match_num","winner_id","loser_id","best_of"))]
```






